---
title: "The cost of admission"
subtitle: "Magic Kingdom Ticket Prices 1971-2018"
author: "Chris Hamm"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
 # pdf_document:
 #   toc: true
 #   number_sections: true
 #   fig_caption: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

# Introduction
Trips to Walt Disney World are expensive and sometimes frustrating. Ever-increasing ticket prices can add to the frustration. It seems that every few months we hear about another hike in ticket prices. With the recent news that [variable pricing](http://time.com/money/5405838/disney-world-ticket-prices-2018/) for tickets is on the way, I wanted to take a look at the history of the cost of a single day admission to the Magic Kingdom and ask a few questions:

1. Are ticket prices rising faster now?
1. Did the different Disney CEOs raise ticket prices at different rates? 
1. Adjusted for inflation, does it cost more to visit Walt Disney World's Magic Kingdom today than it did in 1971? 

### TL;DR

1. Yes, the admission price under Bob Iger is increasing faster than it did under Michael Eisner.
1. No, we can't tell the CEOs apart but this is due to statistical issues with the number of increases made under each CEO. 
1. **Yes**. **Yes**. **Yes**. Adjusted for inflation, the cost of a one-day ticket to the Magic Kingdom in 1971 was \$21.65 and today it is approximately \$122.


# Data

To ask these questions I needed data. I needed data on: the historic price of admission to the Magic Kingdom; the dates those ticket prices were in effect; and the Disney CEO at the time. It wasn't hard to find websites with the history of admission prices. Whenever possible, I selected the most expensive ticket available that was not a park-hopper. I used the data from this [site](http://allears.net/walt-disney-world/wdw-planning/wdw-ticket-increase-guide/) and cross referenced it [here](https://www.travelandleisure.com/trip-ideas/disney-vacations/disney-world-ticket-costs-over-time). I got the dates for the Disney CEOs from the book [Disney War](https://en.wikipedia.org/wiki/DisneyWar) by James B. Stewart and cross referenced them with this [Wikipedia article](https://en.wikipedia.org/wiki/List_of_management_of_The_Walt_Disney_Company). Please let me know if I inadvertently included the incorrect data in this post, and please point me to your source so I can update the post.

Because this is a data science blog post, I've posted the code to a [repository](https://github.com/butterflyology/Theme_Park_prices) so that anyone with a computer and the desire can recreate the analyses presented here.


```{r prelims, echo = FALSE, results = FALSE, message = FALSE}
# Take care of some preliminaries, set the random number seed and load libraries we will use.
set.seed(982352)
options("getSymbols.warning4.0" = FALSE)

library("tidyverse") # This is a "meta-package" that contains lots of other useful packages, like ggplot2 and readr.
library('kableExtra') # We will use this to generate a nice looking table in `html`.
library("lubridate") # we will need some functions from here to work with date data
library("quantmod") # we will use this to get consumer price index data
library("magicfor") # package to save output from a for loop
library("gridExtra") # To plot multiple plots in one pane
```


Let's take a look at the data set I created. There are three columns in the data set:

1. `Date` - The date the admission price was in effect.
1. `Price` - The cost of single-day admission to the Magic Kingdom
1. `CEO` - The Disney CEO at the time of the price.

```{r import, echo = FALSE}
data <- read_csv("../data/MK_prices.csv", col_names = TRUE, col_types = cols(Date = col_date("%Y-%m-%d"), Price = col_double(), CEO = col_character()))

col_names <- c("Date", "Price (USD)", "CEO")
knitr::kable(x = data, format = "html", align = "c", col.names = col_names) %>%
  row_spec(row = 1:2, align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  scroll_box(width = "650px", height = "300px")
```


# Visuals

You always want to look at the data. Here's a plot of admission price by date on the horizontal (x-axis) and the cost of admission to the Magic Kingdom on the vertical axis (y-axis) in US Dollars. Each point's color represents the Disney CEO at the time the price increased.

```{r plot_1, echo = FALSE, results = FALSE}
prices <- data %>% 
  ggplot(aes(x = Date, y = Price, col = CEO)) +
  geom_point(size = 2.5) +
  scale_color_discrete(breaks = c("Disney", "Tatum", "Walker", "Miller", "Eisner", "Iger")) +
  theme_bw() + 
  theme(legend.position = c(0.15, 0.75), legend.text = element_text(size = 15)) +
  labs(y = "Price (USD)") + 
  coord_cartesian(ylim = c(0, 130))
prices
```

Does it look like Mr. Iger has upped the rate of admission increases faster than the other CEOs? Maybe. But this isn't an apples-to-apples comparison. We need to adjust the prices for inflation. We will adjust inflation using the "Consumer Price Index for All Urban Consumers: All Items" from the rock stars over at the Bureau of Labor Statistics.

We will pull out the Consumer Price Index (CPI) for the dates the admission prices went into effect and normalize them based on a 1971 baseline (the year WDW opened). Then we convert the older prices to 2018 US Dollars.

We've added three new columns to out data set:

1. `Year` - We used this column to calculate the CPI
1. `Adjustment` - The adjusted CPI value
1. `Adjusted Price` - The original `Price` of a ticket times the `Adjustment`.

```{r CPI, warning = FALSE, echo = FALSE, results = "hide"}
getSymbols("CPIAUCSL", src = "FRED")
str(CPIAUCSL)

avg.cpi <- apply.yearly(CPIAUCSL, mean)
avg_cpi <- data.frame(Date = index(avg.cpi), avg.cpi)
rownames(avg_cpi) <- NULL
avg_cpi <- avg_cpi %>% 
  mutate(Year = year(Date)) %>%
  select(-Date)

# Inelegant, but it works, add a column with the year to our data set, we will join by this column
new_data <- data %>% 
  mutate(Year = year(Date))
magic_for(print, silent = TRUE)
for(year in new_data$Year){
  print(tail(avg_cpi$CPIAUCSL / avg_cpi[avg_cpi$Year == year, 1], 1))
}
adjustment <- magic_result_as_dataframe() 
colnames(adjustment) <- c("Year", "ADJ")

newer_data <- left_join(x = new_data, y = adjustment, by = "Year")
newer_data <- newer_data %>% 
  mutate(ADJ_Price = Price * ADJ) %>%
  filter(!duplicated(Date))
```

```{r newer_data, echo = FALSE}
newer_col_names <- c("Date", "Price (USD)", "CEO", "Year", "Adjustment", "Adjusted Price (USD)" )
knitr::kable(x = newer_data, format = "html", align = "c", col.names = newer_col_names) %>%
  row_spec(row = 1:2, align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  scroll_box(width = "650px", height = "300px")
```


Now we have our admission prices adjusted for inflation. Let's plot:

```{r adj_plot, echo = FALSE}
newer <- newer_data %>% 
  ggplot(aes(x = Date, y = ADJ_Price, col = CEO)) +
  geom_point(size = 2.5) +
  scale_color_discrete(breaks = c("Disney", "Tatum", "Walker", "Miller", "Eisner", "Iger")) +
  theme_bw() + 
  theme(legend.position = c(0.15, 0.75), legend.text = element_text(size = 15)) +
  labs(y = "Adjusted Price (USD)") +
  coord_cartesian(ylim = c(0, 130))
newer

# This pulls out data from the plot so we can see what color went with each point. 
#ggplot_build(newer)$data
```


It's helpful to see the adjusted prices side by side with the original prices. 

```{r side_by_side, echo = FALSE}
prices <- prices + 
  theme(legend.position = c(0.25, 0.75), legend.text = element_text(size = 15))
newer <- newer +
  theme(legend.position = c(0.25, 0.75), legend.text = element_text(size = 15))

grid.arrange(prices, newer, ncol = 2)
```


What kind of blows my mind is that the Magic Kingdom's prices have, on occasion, not kept up with inflation. When I saw this I manually checked the adjustments I calculated using the Bureau of Labor Statistics own [calculator](https://data.bls.gov/cgi-bin/cpicalc.pl). My calculations were within \$0.20 each time.

On to the questions.


# Eisner vs. Iger

The first two questions are related. To refresh your recollection: 

1. Are ticket prices rising faster now?

Let's look at the rate of increase for each CEO. We will fit a straight line through each CEOs prices and use that line to describe each pricing regime. We will use the same plot as before and overlay lines on to it. Roy O. Disney was CEO for only one price point and won't get a line. 


```{r lines, echo = FALSE, warning = FALSE}
newer_data %>% 
  ggplot(aes(x = Date, y = ADJ_Price, col = CEO)) +
  geom_point(size = 2.5) +
  scale_color_discrete(breaks = c("Disney", "Tatum", "Walker", "Miller", "Eisner", "Iger")) +
  theme_bw() + 
  theme(legend.position = c(0.15, 0.75), legend.text = element_text(size = 15)) +
  labs(y = "Adjusted Price (USD)") +
  coord_cartesian(ylim = c(0, 130)) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, show.legend = FALSE, level = 0.95)
```


The shaded regions around the lines represent the confidence interval around the line. A confidence interval is how sure we are of our estimate of the line, and narrow intervals are better.

But we are interested in the steepness (slope) of the lines.Are the slopes between Michael Eisner and Bob Iger the same? They don't appear to be the same but I want to check. I ran another analysis to ask this question.


```{r ancova, echo = FALSE, results = FALSE}
# create a data set that contains just the Eisner and Iger rows
Eisner_Iger <- newer_data %>% 
  filter(CEO == "Eisner" | CEO == "Iger")

# a model with the interaction term
mod1 <- aov(ADJ_Price ~ Date * CEO, data = Eisner_Iger)
summary(mod1)

# a model without the interaction term
mod2 <- aov(ADJ_Price ~ Date + CEO, data = Eisner_Iger)
summary(mod2)

anova(mod1, mod2)
# Removing the interaction term shows that CEO does significantly affect the model. But we don't know how the slope changes.


# We need to further parse the data and run a linear regression for the two most recent CEOs
Eisner <- newer_data %>%
  filter(CEO == "Eisner")

Iger <- newer_data %>%
  filter(CEO == "Iger")

lm_Eisner <- lm(ADJ_Price ~ Date, data = Eisner)
lm_Iger <- lm(ADJ_Price ~ Date, data = Iger)

summary(lm_Eisner)
summary(lm_Iger)
# The slopes are very different indeed.


# Manually change colors to match previous plots
Eisner_Iger %>%
  ggplot(aes(x = Date, y = ADJ_Price, col = CEO)) +
  geom_point(size = 2.5) +
  labs(y = "Adjusted Price (USD)") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, show.legend = FALSE, level = 0.95) +
  theme_bw() +
  theme(legend.position = c(0.15, 0.75), legend.text = element_text(size = 15)) +
  scale_color_manual(values = c("#B79F00", "#00BA38"))
```


# Rate of increase

Did the different Disney CEOs raise ticket prices at different rates? 

No, we can't tell them apart. But wait, you just said Eisner and Iger were different. If we just look at Eisner and Iger at the same time we can tell them apart, but if look at all the CEOs at the same time we encounter an issue. We have unbalanced data and that limits our ability to make inferences and tell things apart.

```{r CEOs, echo = FALSE, results = "hide", warning = FALSE}
lm1 <- lm(ADJ_Price ~ Date + CEO, data = newer_data)
summary(lm1)


lm2 <- aov(ADJ_Price ~ Date + CEO, data = newer_data)
summary(lm2)
TukeyHSD(lm2, which = 'CEO')
```


# Tracking inflation
Adjusted for inflation, does it cost more to visit Disney today than it did in 1971? 

**Yes**. **Yes**. **Yes**. Adjusted for inflation, the cost of admission to the Magic Kingdom in 1971 was \$21.64. The cost of admission on October 13th, 2018 is \$122.

If the cost of Magic Kingdom admission had simply tracked inflation, the cost today would be \$21.64.

Let's add to our previous plot.

```{r track, echo = FALSE}
yr_1971 <- avg.cpi / as.numeric(avg.cpi['1971']) # using 1971 as the base year
adj_1971 <- data.frame(Date = index(yr_1971), yr_1971)
rownames(adj_1971) <- NULL

adj_1971 <- adj_1971 %>% 
  filter(Date >= "1971-12-01") %>% 
  mutate(adj_1971 = 3.5 * CPIAUCSL, Year = year(Date)) %>% 
  select(-CPIAUCSL)
#dim(adj_1971)

ROD <- left_join(newer_data, adj_1971, by = "Year") %>% 
  select(-Date.y) %>% 
  rename(Date = Date.x)
#dim(ROD)

ROD_plot <- ROD %>% 
  ggplot(aes(y = ADJ_Price, x = Date, col = CEO)) +
  geom_point(size = 2.5) +
  scale_color_discrete(breaks = c("Disney", "Tatum", "Walker", "Miller", "Eisner", "Iger")) +
  theme_bw() +
  theme(legend.position = c(0.15, 0.75), legend.text = element_text(size = 15)) +
  labs(y = "Adjusted Price (USD)") +
  coord_cartesian(ylim = c(0, 130)) +
  geom_point(data = ROD, aes(x = Date, y = adj_1971), col = "red", size = 2.5)

ROD_plot
```

The line of red dots along the bottom of the plot represents the cost of admission to the Magic Kingdom if the cost had just tracked inflation.

